---
title: "ENMPA Homework Assignment Key"
format: 
  html: 
    code-fold: show
    toc: true
    number-sections: true
    embed-resources: true
    message: false 
    warning: false
editor: visual
theme: lumen
execute:
  echo: true
  message: false
  warning: false
---

# library

Use the library enmpa for the tools Use terra for the mapping

```{r}
library(enmpa) #this has the functions for running the models
library(terra) #this package helps with the visulization
library(tidyverse) #keeps things clean
library(here) #makes saving and uploading data sets easier
library(geodata) #has the 
library(predicts)

```
## Getting the Bio Data

This code gets the bioclim data

```{r}
#load bioclim data. This data is used for fitting the model and making a pretty graph
bioclim_data <- worldclim_global(var = "bio",
                                    res = 2.5,
                                    path = "Data/") 
```

## Load, Clean, and Create Psudo Absence Data
**Plot the monarch data. What are the differences and similarities between the monarch map and the milkweed map that we made in the demo?**


```{r}
obs_data_m <- read_csv(here('Data','Monarch.csv')) |>
  select(gbifID, decimalLatitude, decimalLongitude) |> drop_na()|>
  rename(latitude=decimalLatitude) |> 
  rename(longitude=decimalLongitude)

# This bit of code allows us to crop the bioclim data for just the section we have for the milkweed.
max_lat <- ceiling(max(obs_data_m$latitude))
min_lat <- floor(min(obs_data_m$latitude))
max_lon <- ceiling(max(obs_data_m$longitude))
min_lon <- floor(min(obs_data_m$longitude))

# Store boundaries in a single extent object
geographic_extent <- ext(x = c(min_lon, max_lon, min_lat, max_lat))

# This makes a pretty map for our milkweed data.
world_map <- world(resolution = 3,
                   path = "data/")
# Crop the map to our area of interest
my_map <- crop(x = world_map, y = geographic_extent)
# Plot the base map
plot(my_map,
     axes = TRUE,
     col = "grey95")
# Add the points for individual observations
points(x = obs_data_m$longitude,
       y = obs_data_m$latitude,
       col = "darkorange",
       pch = 20,
       cex = 0.75)
#prepare the data for modeling

sample_extent <- geographic_extent * 1.25
# Crop bioclim data to desired extent
bioclim_data <- crop(x = bioclim_data, y = sample_extent)

# Plot the first of the bioclim variables to check on cropping
plot(bioclim_data[[1]]) #this a good way to make sure the boundries are good!

##### Psudo absinces 
set.seed(1234) #This insures we get the same results every time. You can always change it to your favorite number

# Randomly sample points (same number as our observed points)
background_m <- spatSample(x = bioclim_data,
                         size = 500, # generate 500 pseudo-absence points
                         values = FALSE, # don't need values
                         na.rm = TRUE, # don't sample from ocean
                         xy = TRUE) # just need coordinates

# Pull out coordinate columns, x (longitude) first, then y (latitude) from
# saguaro data
presence_m  <- obs_data_m[, c("longitude", "latitude")]

# Add column indicating presence
presence_m$pa <- 1

# Convert background data to a data frame
absence_m <- as.data.frame(background_m)
# Update column names so they match presence points
colnames(absence_m) <- c("longitude", "latitude")
# Add column indicating absence
absence_m$pa <- 0
# Join data into single data frame
all_points_m <- rbind(presence_m, absence_m)


#add the climate data to the milk weed data
bioclim_extract_m <- terra::extract(x = bioclim_data,
                           y = all_points_m[, c("longitude", "latitude")],
                           ID = FALSE) # No need for an ID column 
#### for what ever reason, you need to specify that the extract function is coming from the terra package so the "terra::" portion is absolutely critical

# Add the point and climate datasets together
points_climate_m <- cbind(all_points_m, bioclim_extract_m) 

#select the columns that you need. 
points_climate_m <- points_climate_m |> select(longitude, latitude, pa, wc2.1_2.5m_bio_1, wc2.1_2.5m_bio_12)
```

## Use the calibration_glm() Function to Identify Best Model

**What is the best model?**

```{r}


get_formulas(dependent = "Sp", independent = c("wc2.1_2.5_bio_1", "wc2.1_2.5_bio_12"),  
             type = "lq", mode = "moderate")

points_climate_m <- as.data.frame(points_climate_m, xy = TRUE)

calibration <- calibration_glm(data = points_climate_m, dependent = "pa",
                               independent = c("wc2.1_2.5m_bio_1", "wc2.1_2.5m_bio_12"),
                               response_type = "lpq",
                               formula_mode = "moderate", 
                               exclude_bimodal = TRUE, 
                               selection_criterion = "TSS",
                               cv_kfolds = 5, 
                               verbose = FALSE)
calibration
#> enmpa-class `enmpa_calibration`:
#> $selected             : Selected models (N = 2)
#> $summary              : A summary of statistics for all models. 
#> $calibration_results  : Results obtained from cross-validation for all models. 
#> $data                 : Data used for calibration. 
#> $partitioned_data     : k-fold indexes (k = 5)
#> $weights              : Use of weights (FALSE)
 
summary(calibration)
```

## Fit and Create Projections for the Selected Model

**What is the model saying for monarchs?**

```{r}

```

## Create a Response Curve for the Two Environmental Parameters

**What are the response curves saying about the two environmental parameters?**

```{r}

```

## Compare the Model Progections and Response Curves of the Milkweed and Monarch Data Sets

**Where would you expect monarchs to be seen due to the presence of milkweed but there were no observations of monarchs? What is a probable explanation for that?**

**Where would you expect milkweed to be seen due to the presence of monarchs but there were no observations of milkweed? What is a probable explanation for that?**

**Are there any other surprises in the models or the data?**




